---
title: "MBiol_Bees_Analysis"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
library(ggplot2)
library(dplyr)
#install.packages("ggh4x")
library(ggh4x)
library(here)
library(tidyverse)
library(emmeans)
library(forcats)
```

# Load in data

```{r}
# Load in data
bees_temps <- read.csv(here("Data/26_04_26_bees_temps_5km.csv"))
```

# Set up

```{r}
# Log transform measurements, make sex a factor, and capitalise (for plots)
bees_temps <- bees_temps %>%
  mutate(
    ITD = intertegular_distance_mm,
    HW  = HW_mm,
    FW = FW_length_mm,
    tibia = tibia_length_mm,
    log_ITD = log(ITD),
    log_HW  = log(HW),
    log_FW = log(FW),
    log_tibia = log(tibia),
    # Sex
    sex = factor(sex),
    sex = fct_recode(sex,
      Female = "female",
      Male   = "male"
    ),
    # Ecology
    ecology = factor(ecology),
    ecology = fct_recode(ecology,
      "Cavity Nester"   = "cavity nester",
      "Excavator"       = "excavator",
      "Kleptoparasite"  = "kleptoparasite"
    )
  )

# Rescale year
bees_temps$year_rescaled <- (bees_temps$year - 1800) / 100
# Scaled to "centuries since 1800" to improve model stability

# Set species order so they are organised by ecology and consistent across plots
species_order <- c(
  "Anthidium manicatum",
  "Megachile centuncularis",
  "Hylaeus communis",
  "Hylaeus hyalinatus",
  "Andrena chrysosceles",
  "Andrena wilkella",
  "Colletes succintus",
  "Lasioglossum fulvicorne",
  "Nomada flava",
  "Nomada goodeniana",
  "Nomada ruficornis",
  "Sphecodes geoffrellus"
)

# Keep order and ensure factor
bees_temps$full_name <- factor(bees_temps$full_name, levels = species_order)

```


```{r}
# Helper for italic species labels on axis
italic_species <- function(x) {
  parse(text = paste0("italic('", x, "')"))
}

# Create italicised labels for facets (helper won't work here)
species_labels <- c(
  "Anthidium manicatum" = "italic('Anthidium manicatum')",
  "Megachile centuncularis" = "italic('Megachile centuncularis')",
  "Hylaeus communis" = "italic('Hylaeus communis')",
  "Hylaeus hyalinatus" = "italic('Hylaeus hyalinatus')",
  "Andrena chrysosceles" = "italic('Andrena chrysosceles')",
  "Andrena wilkella" = "italic('Andrena wilkella')",
  "Colletes succintus" = "italic('Colletes succintus')",
  "Lasioglossum fulvicorne" = "italic('Lasioglossum fulvicorne')",
  "Nomada flava" = "italic('Nomada flava')",
  "Nomada goodeniana" = "italic('Nomada goodeniana')",
  "Nomada ruficornis" = "italic('Nomada ruficornis')",
  "Sphecodes geoffrellus" = "italic('Sphecodes geoffrellus')"
)
```


# Data Exploration

## Plots for overall dataset

```{r}
# Plot the number of bee specimens measured from each year
ggplot(bees_temps, aes(x = year)) + 
  geom_bar() + 
  labs(
    title = "Coverage of Specimen Collection Year",
    x = "Year Collected",
    y = "Number of Specimens Measured"
  ) +
  theme(plot.title = element_text(hjust = 0.5))


# Plot the number of specimens of each sex for each species
ggplot(bees_temps, aes(x = full_name, fill = sex)) +
  geom_bar(position = "dodge") +
  scale_x_discrete(labels = italic_species) +
  labs(
    title = "Number of Specimens by Sex per Species",
    x = "Species",
    y = "Number of Specimens Measured",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Per species plots

##### Function for year plot

```{r}
# Function for year plot (set scale for x-axis)
make_species_plot_year_centered <- function(df, xvar, yvar, x_span, y_span, year_center = 1.5) {
  species_levels <- levels(factor(df$full_name))
  x_scales <- lapply(species_levels, function(sp) {
    d <- df %>% filter(full_name == sp)
    if (xvar == "year_rescaled") {
      xmid <- year_center
    } else {
      xmid <- mean(d[[xvar]], na.rm = TRUE)
    }
    xmin <- xmid - x_span / 2
    xmax <- xmid + x_span / 2
    if (xvar == "year_rescaled") {
      scale_x_continuous(
        limits = c(xmin, xmax),
        breaks = seq(
          floor(xmin / 0.5) * 0.5,
          ceiling(xmax / 0.5) * 0.5,
          by = 0.5
        ),
        labels = function(x) round(x * 100 + 1800)
      )
    } else {
      scale_x_continuous(
        limits = c(xmin, xmax)
      )
    }
  })
  
  y_scales <- lapply(species_levels, function(sp) {
    d <- df %>% filter(full_name == sp)
    ymid <- mean(d[[yvar]], na.rm = TRUE)
    ymin <- ymid - y_span / 2
    ymax <- ymid + y_span / 2
    scale_y_continuous(
      limits = c(ymin, ymax),
      breaks = seq(
        floor(ymin / 0.2) * 0.2,
        ceiling(ymax / 0.2) * 0.2,
        by = 0.2
      )
    )
  })
  
  ggplot(df, aes(x = .data[[xvar]], y = .data[[yvar]], colour = sex)) +
    geom_point(alpha = 0.6, size = 1.5) +
    geom_smooth(method = "lm", se = FALSE) +
    facet_wrap(~ full_name,scales = "free", labeller = as_labeller(species_labels, label_parsed)) +
    ggh4x::facetted_pos_scales(
      x = x_scales,
      y = y_scales
    ) +
    labs(
      x = xvar,
      y = yvar,
      colour = "Sex"
    ) +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(hjust = 0.5))
}
```


#### Year and ITD plot

```{r}
# Plot for Year and ITD
make_species_plot_year_centered(
  df = bees_temps,
  xvar = "year_rescaled",
  yvar = "log_ITD",
  x_span = 1.5,
  y_span = 0.8
) +
  labs(
    title = "Effect of Year on Body Size",
    x = "Year",
    y = "log(ITD)"
  )

```

##### Function for temp plot

```{r}
# Function for plots where I can set the window size:
make_species_plot <- function(df, xvar, yvar, x_span, y_span) {
  species_levels <- levels(factor(df$full_name))
  x_scales <- lapply(species_levels, function(sp) {
    d <- df %>% filter(full_name == sp)
    xmid <- mean(d[[xvar]], na.rm = TRUE)
    scale_x_continuous(limits = c(xmid - x_span / 2, xmid + x_span / 2))
  })
  y_scales <- lapply(species_levels, function(sp) {
    d <- df %>% filter(full_name == sp)
    ymid <- mean(d[[yvar]], na.rm = TRUE)
    ymin <- ymid - y_span / 2
    ymax <- ymid + y_span / 2
    scale_y_continuous(
      limits = c(ymin, ymax),
      breaks = seq(
        floor(ymin / 0.2) * 0.2,
        ceiling(ymax / 0.2) * 0.2,
        by = 0.2
      )
    )
  })
  
  ggplot(df, aes(x = .data[[xvar]], y = .data[[yvar]], colour = sex)) +
    geom_point(alpha = 0.6, size = 1.5) +
    geom_smooth(method = "lm", se = FALSE) +
    facet_wrap(~ full_name,scales = "free", labeller = as_labeller(species_labels, label_parsed)) +
    ggh4x::facetted_pos_scales(
      x = x_scales,
      y = y_scales
    ) +
    labs(
      x = xvar,
      y = yvar,
      colour = "Sex"
    ) +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(hjust = 0.5))
}
```

#### Mean temp and ITD plot

```{r}
# Mean temp and ITD
make_species_plot(
  df = bees_temps,
  xvar = "mean_preflight_temp",
  yvar = "log_ITD",
  x_span = 6,
  y_span = 0.7
)+
  labs(
    title = "Effect of temperature on body size",
    x = "Mean pre-flight temperature (°C)",
    y = "log(ITD)"
  )
```

## Summary statistics

###### Calculate the mean and variation in size per species

#### Just log(ITD)

```{r}
size_summary <- bees_temps %>%
  group_by(full_name) %>%
  summarise(
    mean_log_ITD = mean(log_ITD, na.rm = TRUE),
    sd_log_ITD   = sd(log_ITD, na.rm = TRUE),
    n = n()
  )

size_summary
```

#### All log(measured traits)

```{r}
size_summary_all <- bees_temps %>%
  group_by(full_name) %>%
  summarise(
    across(
      c(log_ITD, log_HW, log_FW, log_tibia),
      list(mean = ~mean(.x, na.rm = TRUE),
           sd   = ~sd(.x, na.rm = TRUE)),
      .names = "{.col}_{.fn}"
    ),
    n = n()
  )

size_summary_all
```

#### Raw measurements

#### ITD only

```{r}
size_summary <- bees_temps %>%
  group_by(full_name) %>%
  summarise(
    mean_ITD = round(mean(ITD, na.rm = TRUE), 3),
    sd_ITD   = round(sd(ITD, na.rm = TRUE), 3),
    n = n()
  )

size_summary
```

#### All measurements

```{r}
size_summary_all <- bees_temps %>%
  group_by(full_name) %>%
  summarise(
    across(
      c(ITD, HW, FW, tibia),
       list(mean = ~round(mean(.x, na.rm = TRUE), 3),
           sd   = ~round(sd(.x, na.rm = TRUE), 3)),
      .names = "{.col}_{.fn}"
    ),
    n = n()
  )

size_summary_all
```

#### Just ITD measurements by sex

```{r}
size_summary <- bees_temps %>%
  group_by(full_name, sex) %>%
  summarise(
    mean_ITD = round(mean(ITD, na.rm = TRUE), 3),
    sd_ITD   = round(sd(ITD, na.rm = TRUE), 3),
    n = n(),
    .groups = "drop"
  )

size_summary
```

#### All measurements by sex

```{r}
size_summary_all <- bees_temps %>%
  group_by(full_name, sex) %>%
  summarise(
    across(
      c(ITD, HW, FW, tibia),
      list(mean = ~round(mean(.x, na.rm = TRUE), 3),
           sd   = ~round(sd(.x, na.rm = TRUE), 3)),
      .names = "{.col}_{.fn}"
    ),
    n = n(),
    .groups = "drop"
  )

size_summary_all
```

# LMs analysis

## Create non-interation and sex-interaction models

```{r}
# Separate LM for each species

# Ensure variables are factors
bees_temps <- bees_temps %>%
  mutate(
    sex = factor(sex),
    ecology = factor(ecology),
    full_name = factor(full_name)
  )

species_list <- levels(bees_temps$full_name)

models_main <- list()
models_interactions <- list()

for (sp in species_list) {
  df_sp <- bees_temps %>% filter(full_name == sp)
  
  # Base model
  m1 <- lm(log_ITD ~ sex + mean_preflight_temp + year_rescaled + latitude, 
           data = df_sp)
  # Interaction model
  m2 <- lm(log_ITD ~ sex*mean_preflight_temp + sex*year_rescaled + latitude, 
           data = df_sp)
  
  models_main[[sp]] <- m1
  models_interactions[[sp]] <- m2
}
```

## Compare models

```{r}
# Compare models to see if interaction is needed using AIC

aic_results <- lapply(species_list, function(sp) {
  AIC(models_main[[sp]], models_interactions[[sp]])
})

names(aic_results) <- species_list
aic_results
```



# Calculate and plot the sex-specific effects of temperature and year

```{r}
#Sex-specific effects of temperature and year on log(ITD) using emtrends from interaction models

# Ecology lookup
species_eco <- bees_temps %>%
  distinct(full_name, ecology) %>%
  arrange(ecology, full_name)

species_list <- species_eco$full_name

# Helper to extract sex-specific slopes for any predictor
extract_trends <- function(model_list, var_name, specs_formula = ~ sex) {
  df <- do.call(rbind, lapply(names(model_list), function(sp) {
    mod <- model_list[[sp]]
    emtrends(
      mod,
      specs = specs_formula,
      var = var_name,
      infer = c(TRUE, TRUE)
    ) %>%
      as.data.frame() %>%
      mutate(species = sp)
  }))
  df %>%
    left_join(species_eco, by = c("species" = "full_name")) %>%
    mutate(
      species = factor(species, levels = species_list),
      ecology = factor(ecology),
    )
}

# Helper to plot the slopes

plot_trends <- function(df, slope_col, xlab) {
  ggplot(df, aes(x = .data[[slope_col]], y = species, color = sex)) +
    geom_vline(xintercept = 0, linetype = 2, linewidth = 0.5, colour = "grey60") +
    geom_errorbar(
      aes(xmin = lower.CL, xmax = upper.CL),
      orientation = "y",
      position = position_dodge(width = 0.55),
      width = 0.18,
      linewidth = 0.7
    ) +
    geom_point(
      position = position_dodge(width = 0.55),
      size = 2.4
    ) +
    scale_y_discrete(
      limits = rev,
      labels = function(x) parse(text = paste0("italic('", x, "')"))
    ) +
    facet_grid2(
      ecology ~ .,
      scales = "free_y",
      space = "free_y",
      strip = strip_themed(
        background_y = elem_list_rect(
          fill = c("#fae8b2", "#cb8e73", "#ccb3d6"),
          colour = c(NA, NA, NA),
          linewidth = c(0, 0, 0)
        )
      )
    ) +
    labs(x = xlab, y = NULL, color = "Sex") +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.major.y = element_blank(),
      strip.text.y = element_text(angle = 270, face = "plain"),
      legend.position = "bottom",
      panel.border = element_rect(colour = "#6b6b6b", fill = NA, linewidth = 0.5),
      strip.background = element_blank()
    )
}

```

### Plot for year

```{r}
# Sex-specific slopes for year
year_sex <- extract_trends(models_interactions, "year_rescaled")
plot_trends(year_sex, "year_rescaled.trend", "Estimated effect of collection year on log(ITD)") +
  labs(title = "Estimated Effect of Collection Year on Bee Body Size") +
  theme(plot.title = element_text(hjust = 0.5))
```

### Plot for temperature

```{r}
# Sex-specific slopes for temperature
temp_sex <- extract_trends(models_interactions, "mean_preflight_temp")
plot_trends(temp_sex, "mean_preflight_temp.trend", "Estimated effect of pre-flight temperature on log(ITD)") +
  labs(title = "Estimated Effect of Increased Pre-flight Temperature on Bee Body Size") +
  theme(plot.title = element_text(hjust = 0.5))
```

# Allometry

```{r}
library(dplyr)
library(broom)
library(ggplot2)

# Centre temperature
bees_temps <- bees_temps %>%
  mutate(
    temp_c = mean_preflight_temp - mean(mean_preflight_temp, na.rm = TRUE)
  )

# Create helper function to fit and extract interaction term
extract_allometry <- function(response_var) {
  models <- bees_temps %>%
    split(.$full_name) %>%
    lapply(function(df_sp) {
      formula <- as.formula(paste(response_var, "~ log_ITD * temp_c"))
      lm(formula, data = df_sp)
    })
  bind_rows(lapply(names(models), function(sp) {
    tidy(models[[sp]]) %>%
      filter(term == "log_ITD:temp_c") %>%
      mutate(species = sp)
  })) %>%
    mutate(trait = response_var)
}

# Run function for each trait
coef_allometry <- bind_rows(
  extract_allometry("log_HW"),
  extract_allometry("log_FW"),
  extract_allometry("log_tibia")
)

# Add ecology, ordering, and effect size
coef_allometry <- coef_allometry %>%
  left_join(species_eco, by = c("species" = "full_name")) %>%
  mutate(
    ecology = factor(ecology),
    species = factor(species, levels = rev(species_order)),
    trait = factor(trait, levels = c("log_HW", "log_FW", "log_tibia")),
    slope_change_2SD = estimate * 2
  )

# Plot (faceted by trait)
ggplot(coef_allometry, aes(x = estimate, y = species, colour = ecology)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point(size = 2) +
  geom_errorbar(
    aes(xmin = estimate - std.error,
        xmax = estimate + std.error),
    orientation = "y",
    width = 0.2,
    colour = "black"
  ) +
  facet_grid2(
    ecology ~ trait,
    scales = "free_y",
    space = "free_y",
    labeller = labeller(
      species = label_parsed,
      trait = c(
      log_HW = "log(HW)",
      log_FW = "log(FW)",
      log_tibia = "log(Tibia)"
      )
    ),
    strip = strip_themed(
      background_y = elem_list_rect(
        fill = c("#fae8b2", "#cb8e73", "#ccb3d6"),
        colour = c(NA, NA, NA),
        linewidth = c(0, 0, 0)
      )
    )
  ) +
  scale_y_discrete(
    labels = function(x) parse(text = paste0("italic('", x, "')"))
  ) +
  theme_minimal() +
  theme(
    panel.border = element_rect(colour = "#6b6b6b", fill = NA, linewidth = 0.5),
    strip.text.y = element_text(angle = 270),
    strip.background = element_blank(),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Estimated Effect of Temperature on Allometric Slope",
    x = "Change in allometric slope with increased temperature",
    y = "Species",
    colour = "Ecology"
  ) +
  scale_color_manual(values = c("#ffcb30", "#cb8e73", "#b28fbf"))


# View effect sizes
coef_allometry %>%
  select(species, ecology, trait, estimate, std.error, slope_change_2SD)

```


# Just year

```{r}

extract_allometry_year <- function(response_var) {
  models <- bees_temps %>%
    split(.$full_name) %>%
    lapply(function(df_sp) {
      formula <- as.formula(paste(response_var, "~ log_ITD * year_rescaled"))
      lm(formula, data = df_sp)
    })
  
  bind_rows(lapply(names(models), function(sp) {
    tidy(models[[sp]]) %>%
      filter(term == "log_ITD:year_rescaled") %>%
      mutate(species = sp)
  })) %>%
    mutate(trait = response_var)
}

coef_allometry_year <- bind_rows(
  extract_allometry_year("log_HW"),
  extract_allometry_year("log_FW"),
  extract_allometry_year("log_tibia")
)


coef_allometry_year <- coef_allometry_year %>%
  left_join(species_eco, by = c("species" = "full_name")) %>%
  mutate(
    ecology = factor(ecology),
    species = factor(species, levels = rev(species_order)),
    trait = factor(trait, levels = c("log_HW", "log_FW", "log_tibia")),
    slope_change_2SD = estimate * 2
  )

ggplot(coef_allometry_year, aes(x = estimate, y = species, colour = ecology)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point(size = 2) +
  geom_errorbar(
    aes(xmin = estimate - std.error,
        xmax = estimate + std.error),
    orientation = "y",
    width = 0.2,
    colour = "black"
  ) +
  facet_grid2(
    ecology ~ trait,
    scales = "free_y",
    space = "free_y",
    labeller = labeller(
      species = label_parsed,
      trait = c(
        log_HW = "log(HW)",
        log_FW = "log(FW)",
        log_tibia = "log(Tibia)"
      )
    )
  ) +
  scale_y_discrete(
    labels = function(x) parse(text = paste0("italic('", x, "')"))
  ) +
  theme_minimal() +
  theme(
    panel.border = element_rect(colour = "#6b6b6b", fill = NA, linewidth = 0.5),
    strip.text.y = element_text(angle = 270),
    strip.background = element_blank(),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Estimated Effect of Year on Allometric Slope",
    x = "Change in allometric slope over time",
    y = "Species"
  ) +
  scale_color_manual(values = c("#ffcb30", "#cb8e73", "#b28fbf"))
```

# Both year and temp in model
```{r}
library(dplyr)
library(broom)
library(ggplot2)
library(ggh4x)

# Centre temperature
bees_temps <- bees_temps %>%
  mutate(
    temp_c = mean_preflight_temp - mean(mean_preflight_temp, na.rm = TRUE)
  )


# Create function to fit models and extract interaction terms
extract_allometry_both <- function(response_var) {
  models <- bees_temps %>%
    split(.$full_name) %>%
    lapply(function(df_sp) {
      formula <- as.formula(
        paste(response_var,
              "~ log_ITD * temp_c + log_ITD * year_rescaled + sex + latitude")
      )
      lm(formula, data = df_sp)
    })
  bind_rows(lapply(names(models), function(sp) {
    tidy(models[[sp]]) %>%
      filter(term %in% c("log_ITD:temp_c", "log_ITD:year_rescaled")) %>%
      mutate(species = sp)
  })) %>%
    mutate(trait = response_var)
}


# Run function for all traits
coef_allometry_both <- bind_rows(
  extract_allometry_both("log_HW"),
  extract_allometry_both("log_FW"),
  extract_allometry_both("log_tibia")
)


# Clean and label
coef_allometry_both <- coef_allometry_both %>%
  left_join(species_eco, by = c("species" = "full_name")) %>%
  mutate(
    ecology = factor(ecology),
    species = factor(species, levels = rev(species_order)),
    trait = factor(trait, levels = c("log_HW", "log_FW", "log_tibia")),
    predictor = term 
  )

# Plot Temperature effect (controlling for year, sex, + latitude)
plot_temp <- coef_allometry_both %>%
  filter(predictor == "log_ITD:temp_c") %>%
  ggplot(aes(x = estimate, y = species, colour = ecology)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point(size = 2) +
  geom_errorbar(
    aes(xmin = estimate - std.error,
        xmax = estimate + std.error),
    orientation = "y",
    width = 0.2,
    colour = "black"
  ) +
  facet_grid2(
    ecology ~ trait,
    scales = "free_y",
    space = "free_y",
    labeller = labeller(
      trait = c(
      log_HW = "log(HW)",
      log_FW = "log(FW)",
      log_tibia = "log(Tibia)"
      )
    ),
    strip = strip_themed(
      background_y = elem_list_rect(
        fill = c("#fae8b2", "#cb8e73", "#ccb3d6"),
        colour = c(NA, NA, NA),
        linewidth = c(0, 0, 0)
      )
    )
  ) +
  scale_y_discrete(
    labels = function(x) parse(text = paste0("italic('", x, "')"))
  ) +
  theme_minimal() +
  theme(
    panel.border = element_rect(colour = "#6b6b6b", fill = NA),
    strip.text.y = element_text(angle = 270),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Effect of Temperature on Allometric Slope",
    x = "Change in allometric slope with temperature",
    y = "Species"
  ) +
  scale_color_manual(values = c("#ffcb30", "#cb8e73", "#b28fbf"))


# 6. Plot Year effect (controlling for temperature, sex, + latitude)
plot_year <- coef_allometry_both %>%
  filter(predictor == "log_ITD:year_rescaled") %>%
  ggplot(aes(x = estimate, y = species, colour = ecology)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point(size = 2) +
  geom_errorbar(
    aes(xmin = estimate - std.error,
        xmax = estimate + std.error),
    orientation = "y",
    width = 0.2,
    colour = "black"
  ) +
  facet_grid2(
    ecology ~ trait,
    scales = "free_y",
    space = "free_y",
    labeller = labeller(
      trait = c(
      log_HW = "log(HW)",
      log_FW = "log(FW)",
      log_tibia = "log(Tibia)"
      )
    ),
    strip = strip_themed(
      background_y = elem_list_rect(
        fill = c("#fae8b2", "#cb8e73", "#ccb3d6"),
        colour = c(NA, NA, NA),
        linewidth = c(0, 0, 0)
      )
    )
  ) +
  scale_y_discrete(
    labels = function(x) parse(text = paste0("italic('", x, "')"))
  ) +
  theme_minimal() +
  theme(
    panel.border = element_rect(colour = "#6b6b6b", fill = NA),
    strip.text.y = element_text(angle = 270),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Effect of Collection Year on Allometric Slope",
    x = "Change in allometric slope over time",
    y = "Species"
  ) +
  scale_color_manual(values = c("#ffcb30", "#cb8e73", "#b28fbf"))

# ----------------------------
# 7. View plots
# ----------------------------
plot_temp
plot_year
```



```{r}
library(dplyr)
library(broom)
library(ggplot2)
library(ggh4x)

# ----------------------------
# Centre temperature
# ----------------------------
bees_temps <- bees_temps %>%
  mutate(
    temp_c = mean_preflight_temp - mean(mean_preflight_temp, na.rm = TRUE)
  )

# ----------------------------
# Function: fit models separately by sex
# ----------------------------
extract_allometry_by_sex <- function(response_var) {
  
  models <- bees_temps %>%
    split(list(.$full_name, .$sex), drop = TRUE) %>%
    lapply(function(df_sp) {
      formula <- as.formula(
        paste(response_var,
              "~ log_ITD * temp_c + log_ITD * year_rescaled + latitude")
      )
      lm(formula, data = df_sp)
    })
  
  bind_rows(lapply(names(models), function(name) {
    
    # Extract species + sex from name
    parts <- strsplit(name, "\\.")[[1]]
    sp <- parts[1]
    sx <- parts[2]
    
    tidy(models[[name]]) %>%
      filter(term %in% c("log_ITD:temp_c", "log_ITD:year_rescaled")) %>%
      mutate(species = sp,
             sex = sx)
  })) %>%
    mutate(trait = response_var)
}

# ----------------------------
# Run for all traits
# ----------------------------
coef_allometry_sex <- bind_rows(
  extract_allometry_by_sex("log_HW"),
  extract_allometry_by_sex("log_FW"),
  extract_allometry_by_sex("log_tibia")
)

# ----------------------------
# Clean + label
# ----------------------------
coef_allometry_sex <- coef_allometry_sex %>%
  left_join(species_eco, by = c("species" = "full_name")) %>%
  mutate(
    ecology = factor(ecology),
    species = factor(species, levels = rev(species_order)),
    trait = factor(trait, levels = c("log_HW", "log_FW", "log_tibia")),
    predictor = term,
    sex = factor(sex, levels = c("Female", "Male"))
  )

# ----------------------------
# Function to make plots by sex
# ----------------------------
make_plot <- function(df, predictor_name, title_text, xlab_text) {
  
  ggplot(df %>% filter(predictor == predictor_name),
         aes(x = estimate, y = species, colour = ecology)) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    geom_point(size = 2) +
    geom_errorbar(
      aes(xmin = estimate - std.error,
          xmax = estimate + std.error),
      orientation = "y",
      width = 0.2,
      colour = "black"
    ) +
    facet_grid2(
      ecology ~ trait,
      scales = "free_y",
      space = "free_y",
      labeller = labeller(
        trait = c(
          log_HW = "log(HW)",
          log_FW = "log(FW)",
          log_tibia = "log(Tibia)"
        )
      ),
      strip = strip_themed(
        background_y = elem_list_rect(
          fill = c("#fae8b2", "#cb8e73", "#ccb3d6"),
          colour = c(NA, NA, NA),
          linewidth = c(0, 0, 0)
        )
      )
    ) +
    scale_y_discrete(
      labels = function(x) parse(text = paste0("italic('", x, "')"))
    ) +
    theme_minimal() +
    theme(
      panel.border = element_rect(colour = "#6b6b6b", fill = NA),
      strip.text.y = element_text(angle = 270),
      legend.position = "none",
      plot.title = element_text(hjust = 0.5)
    ) +
    labs(
      title = title_text,
      x = xlab_text,
      y = "Species"
    ) +
    scale_color_manual(values = c("#ffcb30", "#cb8e73", "#b28fbf"))
}

# ----------------------------
# Split data by sex
# ----------------------------
coef_female <- coef_allometry_sex %>% filter(sex == "Female")
coef_male   <- coef_allometry_sex %>% filter(sex == "Male")

# ----------------------------
# Temperature plots
# ----------------------------
plot_temp_female <- make_plot(
  coef_female,
  "log_ITD:temp_c",
  "Effect of Temperature on Allometric Slope (Females)",
  "Change in allometric slope with temperature"
)

plot_temp_male <- make_plot(
  coef_male,
  "log_ITD:temp_c",
  "Effect of Temperature on Allometric Slope (Males)",
  "Change in allometric slope with temperature"
)

# ----------------------------
# Year plots
# ----------------------------
plot_year_female <- make_plot(
  coef_female,
  "log_ITD:year_rescaled",
  "Effect of Collection Year on Allometric Slope (Females)",
  "Change in allometric slope over time"
)

plot_year_male <- make_plot(
  coef_male,
  "log_ITD:year_rescaled",
  "Effect of Collection Year on Allometric Slope (Males)",
  "Change in allometric slope over time"
)

# ----------------------------
# View plots
# ----------------------------
plot_temp_female
plot_temp_male

plot_year_female
plot_year_male
```









