---
title: "MBiol_Bees_Analysis"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
library(ggplot2)
library(dplyr)
#install.packages("ggh4x")
library(ggh4x)
library(here)
library(tidyverse)
library(emmeans)
library(forcats)
library(broom)
```

# Load in data

```{r}
# Load in data
bees_temps <- read.csv(here("Data/26_04_26_bees_temps_5km.csv"))
```

# Set up

```{r}
# Log transform measurements, make sex a factor, and capitalise (for plots)
bees_temps <- bees_temps %>%
  mutate(
    ITD = intertegular_distance_mm,
    HW  = HW_mm,
    FW = FW_length_mm,
    TIB = tibia_length_mm,
    log_ITD = log(ITD),
    log_HW  = log(HW),
    log_FW = log(FW),
    log_TIB = log(TIB),
    # Sex
    sex = factor(sex),
    sex = fct_recode(sex,
      Female = "female",
      Male   = "male"
    ),
    # Ecology
    ecology = factor(ecology),
    ecology = fct_recode(ecology,
      "Cavity Nester"   = "cavity nester",
      "Excavator"       = "excavator",
      "Kleptoparasite"  = "kleptoparasite"
    )
  )

# Rescale year
bees_temps$year_rescaled <- (bees_temps$year - 1800) / 100
# Scaled to "centuries since 1800" to improve model stability

# Set species order so they are organised by ecology and consistent across plots
species_order <- c(
  "Anthidium manicatum",
  "Megachile centuncularis",
  "Hylaeus communis",
  "Hylaeus hyalinatus",
  "Andrena chrysosceles",
  "Andrena wilkella",
  "Colletes succintus",
  "Lasioglossum fulvicorne",
  "Nomada flava",
  "Nomada goodeniana",
  "Nomada ruficornis",
  "Sphecodes geoffrellus"
)

# Keep order and ensure factor
bees_temps$full_name <- factor(bees_temps$full_name, levels = species_order)

```

### Set up for visual elements in plots

```{r}
# Helper for italic species labels on axis
italic_species <- function(x) {
  parse(text = paste0("italic('", x, "')"))
}

# Create italicised labels for facets (helper won't work here)
species_labels <- c(
  "Anthidium manicatum" = "italic('Anthidium manicatum')",
  "Megachile centuncularis" = "italic('Megachile centuncularis')",
  "Hylaeus communis" = "italic('Hylaeus communis')",
  "Hylaeus hyalinatus" = "italic('Hylaeus hyalinatus')",
  "Andrena chrysosceles" = "italic('Andrena chrysosceles')",
  "Andrena wilkella" = "italic('Andrena wilkella')",
  "Colletes succintus" = "italic('Colletes succintus')",
  "Lasioglossum fulvicorne" = "italic('Lasioglossum fulvicorne')",
  "Nomada flava" = "italic('Nomada flava')",
  "Nomada goodeniana" = "italic('Nomada goodeniana')",
  "Nomada ruficornis" = "italic('Nomada ruficornis')",
  "Sphecodes geoffrellus" = "italic('Sphecodes geoffrellus')"
)
```

# Data Exploration

## Plots for overall dataset

```{r}
# Plot the number of bee specimens measured from each year
ggplot(bees_temps, aes(x = year)) + 
  geom_bar() + 
  labs(
#    title = "Coverage of Specimen Collection Year",
    x = "Year Collected",
    y = "Number of Specimens Measured"
  ) #+
#  theme(plot.title = element_text(hjust = 0.5))


# Plot the number of specimens of each sex for each species
ggplot(bees_temps, aes(x = full_name, fill = sex)) +
  geom_bar(position = "dodge") +
  scale_x_discrete(labels = italic_species) +
  labs(
#    title = "Number of Specimens by Sex per Species",
    x = "Species",
    y = "Number of Specimens Measured",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Calculate the average number of specimens per species

species_counts <- bees_temps %>%
  count(full_name)

mean(species_counts$n)

```


```{r}
# Plot each species with temp against year 

# Strip colours in the same order as the facets
strip_cols <- bees_temps |>
  distinct(full_name, ecology) |>
  arrange(full_name) |>
  mutate(
    strip_colour = case_when(
      ecology == "Cavity Nester" ~ "#fae8b2",
      ecology == "Excavator" ~ "#cb8e73",
      ecology == "Kleptoparasite" ~ "#ccb3d6"
    )
  ) |>
  pull(strip_colour)


# Dummy data used only to create the legend
legend_data <- data.frame(
  ecology = c("Cavity Nester", "Excavator", "Kleptoparasite"),
  x = min(bees_temps$year, na.rm = TRUE),
  y = min(bees_temps$mean_preflight_temp, na.rm = TRUE)
)


# Centre each panel on the species mean temperature
sp_limits <- bees_temps %>%
  group_by(full_name) %>%
  summarise(
    centre = mean(mean_preflight_temp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(full_name)

# 6 degrees total span = 3 degrees either side of the centre
y_scales <- lapply(sp_limits$centre, function(cen) {
  scale_y_continuous(limits = c(cen - 3, cen + 3))
})

ggplot(bees_temps, 
       aes(x = year, 
           y = mean_preflight_temp)) +
  geom_point(alpha = 0.35, colour = "#797d80") +
  geom_smooth(method = "lm", se = FALSE, colour = "#585b5e", linewidth = 0.7) +
  geom_point(
    data = legend_data,
    aes(x = x, y = y, fill = ecology),
    shape = 22,
    size = 5,
    colour = NA,
    inherit.aes = FALSE,
    na.rm = TRUE
  ) +
  guides(
    fill = guide_legend(
      override.aes = list(colour = NA)
    )
  ) +
  facet_wrap2(
    ~ full_name,
    scales = "free_y",
    strip = strip_themed(
      background_x = elem_list_rect(fill = strip_cols, color = rep("transparent", 12)
                                    )
    )
  ) +
  facetted_pos_scales(y = y_scales) +
  scale_fill_manual(
    values = c(
      "Cavity Nester" = "#fae8b2",
      "Excavator" = "#cb8e73",
      "Kleptoparasite" = "#ccb3d6"
    ),
    name = "Nesting Ecology"
  ) +
  labs(
    x = "Collection year",
    y = "Mean pre-flight temperature (ºC)"
  ) +
  theme_bw() +
  theme(
    panel.border = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(face = "italic"),
    legend.position = "right"
  )
```



```{r}
# Correlation for year and temperature by species
bees_temps %>%
  group_by(full_name) %>%
  summarise(
    n = sum(complete.cases(year, mean_preflight_temp)),
    cor_year_temp = cor(year, mean_preflight_temp, use = "complete.obs")
  )

# Correlation for latitude and temperature by species
bees_temps %>%
  group_by(full_name) %>%
  summarise(
    n = sum(complete.cases (latitude, mean_preflight_temp)),
    cor_temp_lat = cor(latitude, mean_preflight_temp, use = "complete.obs")
  )

```



## Per species plots

#### Function for year plot

```{r}
# Function for year plot (set scale for x-axis)
make_species_plot_year_centered <- function(df, xvar, yvar, x_span, y_span, year_center = 1.5) {
  species_levels <- levels(factor(df$full_name))
  x_scales <- lapply(species_levels, function(sp) {
    d <- df %>% filter(full_name == sp)
    if (xvar == "year_rescaled") {
      xmid <- year_center
    } else {
      xmid <- mean(d[[xvar]], na.rm = TRUE)
    }
    xmin <- xmid - x_span / 2
    xmax <- xmid + x_span / 2
    if (xvar == "year_rescaled") {
      scale_x_continuous(
        limits = c(xmin, xmax),
        breaks = seq(
          floor(xmin / 0.5) * 0.5,
          ceiling(xmax / 0.5) * 0.5,
          by = 0.5
        ),
        labels = function(x) round(x * 100 + 1800)
      )
    } else {
      scale_x_continuous(
        limits = c(xmin, xmax)
      )
    }
  })
  
  y_scales <- lapply(species_levels, function(sp) {
    d <- df %>% filter(full_name == sp)
    ymid <- mean(d[[yvar]], na.rm = TRUE)
    ymin <- ymid - y_span / 2
    ymax <- ymid + y_span / 2
    scale_y_continuous(
      limits = c(ymin, ymax),
      breaks = seq(
        floor(ymin / 0.2) * 0.2,
        ceiling(ymax / 0.2) * 0.2,
        by = 0.2
      )
    )
  })
  
  ggplot(df, aes(x = .data[[xvar]], y = .data[[yvar]], colour = sex)) +
    geom_point(alpha = 0.6, size = 1.5) +
    geom_smooth(method = "lm", se = FALSE) +
    facet_wrap(~ full_name,scales = "free", labeller = as_labeller(species_labels, label_parsed)) +
    ggh4x::facetted_pos_scales(
      x = x_scales,
      y = y_scales
    ) +
    labs(
      x = xvar,
      y = yvar,
      colour = "Sex"
    ) +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(hjust = 0.5))
}
```

#### Year and ITD plot

```{r}
# Plot for Year and ITD
make_species_plot_year_centered(
  df = bees_temps,
  xvar = "year_rescaled",
  yvar = "log_ITD",
  x_span = 1.5,
  y_span = 0.8
) +
  labs(
#    title = "Effect of Year on Body Size",
    x = "Year",
    y = "log(ITD)"
  ) +
  theme(
    legend.position = "right"
  )

```

#### Function for temp plot

```{r}
# Function for plots where I can set the window size:
make_species_plot <- function(df, xvar, yvar, x_span, y_span) {
  species_levels <- levels(factor(df$full_name))
  x_scales <- lapply(species_levels, function(sp) {
    d <- df %>% filter(full_name == sp)
    xmid <- mean(d[[xvar]], na.rm = TRUE)
    scale_x_continuous(limits = c(xmid - x_span / 2, xmid + x_span / 2))
  })
  y_scales <- lapply(species_levels, function(sp) {
    d <- df %>% filter(full_name == sp)
    ymid <- mean(d[[yvar]], na.rm = TRUE)
    ymin <- ymid - y_span / 2
    ymax <- ymid + y_span / 2
    scale_y_continuous(
      limits = c(ymin, ymax),
      breaks = seq(
        floor(ymin / 0.2) * 0.2,
        ceiling(ymax / 0.2) * 0.2,
        by = 0.2
      )
    )
  })
  
  ggplot(df, aes(x = .data[[xvar]], y = .data[[yvar]], colour = sex)) +
    geom_point(alpha = 0.6, size = 1.5) +
    geom_smooth(method = "lm", se = FALSE) +
    facet_wrap(~ full_name,scales = "free", labeller = as_labeller(species_labels, label_parsed)) +
    ggh4x::facetted_pos_scales(
      x = x_scales,
      y = y_scales
    ) +
    labs(
      x = xvar,
      y = yvar,
      colour = "Sex"
    ) +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(hjust = 0.5))
}
```

#### Mean temp and ITD plot

```{r}
# Mean temp and ITD
make_species_plot(
  df = bees_temps,
  xvar = "mean_preflight_temp",
  yvar = "log_ITD",
  x_span = 6,
  y_span = 0.7
)+
  labs(
#    title = "Effect of temperature on body size",
    x = "Mean pre-flight temperature (°C)",
    y = "log(ITD)"
  )+
  theme(
    legend.position = "right"
  )
```

## Summary statistics

###Calculate the mean and variation in size per species

#### ITD only

```{r}
size_summary <- bees_temps %>%
  group_by(full_name) %>%
  summarise(
    mean_ITD = round(mean(ITD, na.rm = TRUE), 3),
    sd_ITD   = round(sd(ITD, na.rm = TRUE), 3),
    n = n()
  )

size_summary
```

#### All measurements by sex

```{r}
size_summary_all <- bees_temps %>%
  group_by(full_name, sex) %>%
  summarise(
    across(
      c(ITD, HW, FW, TIB),
      list(mean = ~round(mean(.x, na.rm = TRUE), 3),
           sd   = ~round(sd(.x, na.rm = TRUE), 3)),
      .names = "{.col}_{.fn}"
    ),
    n = n(),
    .groups = "drop"
  )

size_summary_all
```

# Multiple linear regression analysis

## Create non-interation and sex-interaction models

```{r}
# Separate MLRs for each species

# Ensure variables are factors
bees_temps <- bees_temps %>%
  mutate(
    sex = factor(sex),
    ecology = factor(ecology),
    full_name = factor(full_name)
  )

# Centre temperature
bees_temps <- bees_temps %>%
  mutate(
    temp_c = mean_preflight_temp - mean(mean_preflight_temp, na.rm = TRUE)
  )

species_list <- levels(bees_temps$full_name)

models_base <- list()
models_interactions <- list()

for (sp in species_list) {
  df_sp <- bees_temps %>% filter(full_name == sp)
  
  # Base model
  m1 <- lm(log_ITD ~ sex + temp_c + year_rescaled + latitude, 
           data = df_sp)
  # Interaction model
  m2 <- lm(log_ITD ~ sex*temp_c + sex*year_rescaled + latitude, 
           data = df_sp)
  
  models_base[[sp]] <- m1
  models_interactions[[sp]] <- m2
}
```

## Compare models

```{r}
# Compare models to see if interaction is needed using AIC

aic_results <- lapply(species_list, function(sp) {
  AIC(models_base[[sp]], models_interactions[[sp]])
})

names(aic_results) <- species_list
aic_results
```

# Calculate and plot the sex-specific effects of temperature and year

```{r}
#Sex-specific effects of temperature and year on log(ITD) using emtrends from interaction models

# Ecology lookup
species_eco <- bees_temps %>%
  distinct(full_name, ecology) %>%
  arrange(ecology, full_name)

species_list <- species_eco$full_name

# Helper to extract sex-specific slopes for any predictor
extract_trends <- function(model_list, var_name, specs_formula = ~ sex) {
  df <- do.call(rbind, lapply(names(model_list), function(sp) {
    mod <- model_list[[sp]]
    emtrends(
      mod,
      specs = specs_formula,
      var = var_name,
      infer = c(TRUE, TRUE)
    ) %>%
      as.data.frame() %>%
      mutate(species = sp)
  }))
  df %>%
    left_join(species_eco, by = c("species" = "full_name")) %>%
    mutate(
      species = factor(species, levels = species_list),
      ecology = factor(ecology),
    )
}

# Helper to plot the slopes

plot_trends <- function(df, slope_col, xlab) {
  ggplot(df, aes(x = .data[[slope_col]], y = species, color = sex)) +
    geom_vline(xintercept = 0, linetype = 2, linewidth = 0.5, colour = "grey60") +
    geom_errorbar(
      aes(xmin = lower.CL, xmax = upper.CL),
      orientation = "y",
      position = position_dodge(width = 0.55),
      width = 0.18,
      linewidth = 0.7
    ) +
    geom_point(
      position = position_dodge(width = 0.55),
      size = 2.4
    ) +
    scale_y_discrete(
      limits = rev,
      labels = function(x) parse(text = paste0("italic('", x, "')"))
    ) +
    facet_grid2(
      ecology ~ .,
      scales = "free_y",
      space = "free_y",
      strip = strip_themed(
        background_y = elem_list_rect(
          fill = c("#fae8b2", "#cb8e73", "#ccb3d6"),
          colour = c(NA, NA, NA),
          linewidth = c(0, 0, 0)
        )
      )
    ) +
    labs(x = xlab, y = NULL, color = "      Sex") +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.major.y = element_blank(),
      strip.text.y = element_text(angle = 270, face = "plain"),
      legend.position = "right",
      panel.border = element_rect(colour = "#6b6b6b", fill = NA, linewidth = 0.5),
      strip.background = element_blank()
    )
}

```

### Plot for year

```{r}
# Sex-specific slopes for year
year_sex <- extract_trends(models_interactions, "year_rescaled")
plot_trends(year_sex, "year_rescaled.trend", "Estimated effect of collection year on log(ITD)") #+
#  labs(title = "Estimated Effect of Collection Year on Bee Body Size") +
#  theme(plot.title = element_text(hjust = 0.5))
```

### Plot for temperature

```{r}
# Sex-specific slopes for temperature
temp_sex <- extract_trends(models_interactions, "temp_c")
plot_trends(temp_sex, "temp_c.trend", "Estimated effect of pre-flight temperature on log(ITD)") #+
#  labs(title = "Estimated Effect of Increased Pre-flight Temperature on Bee Body Size") +
#  theme(plot.title = element_text(hjust = 0.5))
```

# Allometry

### Separate sexes and custom x axis

```{r}
# Fit the allometry models separately by species and sex
extract_allometry_by_sex <- function(response_var) {
  models <- bees_temps %>%
    split(list(.$full_name, .$sex), drop = TRUE) %>%
    lapply(function(df_sp) {
      lm(
        as.formula(
          paste(response_var, "~ log_ITD * temp_c + log_ITD * year_rescaled + latitude")
        ),
        data = df_sp
      )
    })
  
  bind_rows(lapply(names(models), function(name) {
    parts <- strsplit(name, "\\.")[[1]]
    sp <- parts[1]
    sx <- parts[2]
    
    tidy(models[[name]]) %>%
      filter(term %in% c("log_ITD:temp_c", "log_ITD:year_rescaled")) %>%
      mutate(
        species = sp,
        sex = sx
      )
  })) %>%
    mutate(trait = response_var)
}

# Run the function for all traits
coef_allometry_sex <- bind_rows(
  extract_allometry_by_sex("log_HW"),
  extract_allometry_by_sex("log_FW"),
  extract_allometry_by_sex("log_TIB")
)


# Clean and label
coef_allometry_sex <- coef_allometry_sex %>%
  left_join(species_eco, by = c("species" = "full_name")) %>%
  mutate(
    ecology = factor(ecology),
    species = factor(species, levels = rev(species_order)),
    trait = factor(trait, levels = c("log_HW", "log_FW", "log_TIB")),
    predictor = term,
    sex = factor(sex, levels = c("Female", "Male"))
  )


# Split into sex-specific data frames
coef_female <- coef_allometry_sex %>% filter(sex == "Female")
coef_male   <- coef_allometry_sex %>% filter(sex == "Male")

# Manually set x-axis limits

# Both fitting but own ranges
#year_xmin <- -1.25
#year_xmax <-  2.42
#temp_xmin <- -0.40
#temp_xmax <-  0.8

# Both same range but cut off some points
#year_xmin <- -1
#year_xmax <-  1
#temp_xmin <- -1
#temp_xmax <-  1

# Both fit, same ranges but squished
#year_xmin <- -1.25
#year_xmax <-  2.42
#temp_xmin <- -1.25
#temp_xmax <-  2.42


# After changing to CI
# fit all
#year_xmin <- -1.9
#year_xmax <-  4
#temp_xmin <- -0.52
#temp_xmax <-  1.2

# Fit all points but not CIs
year_xmin <- -1.9
year_xmax <-  1.9
temp_xmin <- -0.52
temp_xmax <-  0.52



# Create plotting function
make_plot <- function(df, predictor_name, xlab_text, x_min, x_max) {
  ggplot(
    df %>% filter(predictor == predictor_name),
    aes(x = estimate, y = species, colour = ecology)
  ) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "#94abb0") +
    geom_point(size = 2) +
    geom_errorbar(
      aes(
        xmin = estimate - 1.96 * std.error,
        xmax = estimate + 1.96 * std.error
      ),
      orientation = "y",
      width = 0.2,
      colour = "black"
    ) +
    coord_cartesian(xlim = c(x_min, x_max)) +
    facet_grid2(
      ecology ~ trait,
      scales = "free_y",
      space = "free_y",
      labeller = labeller(
        trait = c(
          log_HW = "log(HW)",
          log_FW = "log(FW)",
          log_TIB = "log(TIB)"
        )
      ),
      strip = strip_themed(
        background_y = elem_list_rect(
          fill = c("#fae8b2", "#cb8e73", "#ccb3d6"),
          colour = c(NA, NA, NA),
          linewidth = c(0, 0, 0)
        )
      )
    ) +
    scale_y_discrete(
      labels = function(x) parse(text = paste0("italic('", x, "')"))
    ) +
    theme_minimal() +
    theme(
      panel.border = element_rect(colour = "#6b6b6b", fill = NA),
      strip.text.y = element_text(angle = 270),
      legend.position = "none",
      plot.title = element_text(hjust = 0.5)
    ) +
    labs(
      x = xlab_text,
      y = "Species"
    ) +
    scale_color_manual(values = c("#ffcb30", "#cb8e73", "#b28fbf"))
}


# Create plots
plot_year_female <- make_plot(
  coef_female,
  "log_ITD:year_rescaled",
  "Change in allometric slope over time",
  year_xmin, year_xmax
)

plot_year_male <- make_plot(
  coef_male,
  "log_ITD:year_rescaled",
  "Change in allometric slope over time",
  year_xmin, year_xmax
)

plot_temp_female <- make_plot(
  coef_female,
  "log_ITD:temp_c",
  "Change in allometric slope with temperature",
  temp_xmin, temp_xmax
)

plot_temp_male <- make_plot(
  coef_male,
  "log_ITD:temp_c",
  "Change in allometric slope with temperature",
  temp_xmin, temp_xmax
)

# View plots
plot_year_female
plot_year_male
plot_temp_female
plot_temp_male
```
